---
title: "DATA472 Assignment 3"
author: 'Loren Whithair (Student ID: 300 434 521)'
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  attr.source = ".numberLines")
```

# Question 1

## Code, with comments to identify randomness

Any line of code that adds randomness has been tagged as `#RANDOM`, with an additional tag of `INDEPENDENT` if the simulated values are independent of each other, and `DEPENDENT` otherwise.

First, initialise the variables for simulation and the population:
```{r q1_init_vars}
set.seed(42)

N <- 50 # population
L <- 100 # genetic markers
num_gen <- 20 # number of generations we will breed for
p <- 0.6 # initial probability of a "0" variant
```

Then, initialise a population as a matrix:
```{r q1_init_pop}
# get enough markers for the whole population
data <- sample(0:1, N*L*2, replace=TRUE, prob=c(p,1-p)) #RANDOM INDEPENDENT

# population matrix (one row per individual, one column per copy of a marker)
df <- matrix(data, nrow=N)
```

Create a function that breeds a new generation from the existing population:
```{r q1_breed_fn}
breed_new_gen <- function(df) {
  df_new <- df # work on a copy, so we always have initial population
  
  for (i in 1:nrow(df)){ # for every individual in the population...
    parents <- sample(1:nrow(df), 2, replace=FALSE) #RANDOM DEPENDENT
    
    for (j in 1:L) {# for every marker copy...
      # pick an index of which copy we will take from parent 1 and parent 2
      allele1 <- sample(1:2, 1, replace=TRUE) #RANDOM INDEPENDENT
      allele2 <- sample(1:2, 1, replace=TRUE) #RANDOM INDEPENDENT
      
      # get the variants from the parents
      copy1 <- df[parents[1], 2*(j-1)+ allele1]
      copy2 <- df[parents[2], 2*(j-1)+ allele2]
      
      # insert the parent's marker copies into the child's
      df_new[i,2*(j-1)+1] <- copy1
      df_new[i,2*(j-1)+2] <- copy2
    }
  }
  df_new
}
```

Using these components to breed for the specified number of generations:
```{r q1_breed}
df_new <- df
all_freqs <- matrix(NA, nrow=num_gen, ncol=L) #store the frequencies

for (g in 1:num_gen) {
  df_new <- breed_new_gen(df_new) # one round of breeding
  
  # save frequencies of this generation in row g. Each column is a 
  all_freqs[g,] <- sapply(1:L, function(j) {
    cols <- 2*(j-1) + 1:2
    sum(df_new[,cols] == 0)/N/2
  })
}
```

We now have `df` as the original generation and `df_new` as the final generation (50th).

```{r q1_sample_freq}

# get indices of individuals to survey from  1st and final population
s <- 20 # sample size
p1_idxs <- sample(1:N, s, replace=FALSE) #RANDOM DEPENDENT
pf_idxs <- sample(1:N, s, replace=FALSE) #RANDOM DEPENDENT

# get the individuals
df1 <- df[p1_idxs,]
dff <- df_new[pf_idxs,]

# calculate frequencies for histogram
freqs1 <- sapply(1:L, function(j) {
  cols <- 2*(j-1) + 1:2
  sum(df1[,cols] == 0)/s/2
})

freqsf <- sapply(1:L, function(j) {
  cols <- 2*(j-1) + 1:2
  sum(dff[,cols] == 0)/s/2
})

# generate histograms
xl <- c(0, 1)
yl <- c(0, 30)
par(mfrow=c(2,1))
hist(freqs1, 
     main="Frequencies of allele 0 in sample of first generation, across markers",
     xlab="Frequency of allele 0",
     ylab="Markers with this frequency",
     xlim=xl, ylim=yl
)
hist(freqsf, 
     main="Frequencies of allele 0 in sample of final generation, across markers",
     xlab="Frequency of allele 0",
     ylab="Markers with this frequency",
     xlim=xl, ylim=yl
)
```

## Observations of frequency

Both histograms are centered around approximately  `0.6`, the initial probability of a "0" variant.

The difference in variance demonstrates **genetic drift** - we can see that for the first generation, the frequency of a "0" for a given marker has 40% minimum, 80% maximum. For the final generation, there are at least some markers that consist entirely of the "0" variant.


# Question 2

```{r q2_init_vars}

set.seed(99)

num_days <- 1000
lambda_morn <- 1/8
lambda_aft<- 1/4

# store the results every day
results <- rep(NA, num_days)
```

```{r q2_sim_customers}

for (i in 1:num_days){
  t <- 0 #time passed in the day
  c <- 0 #customers
  while(t < 180){ # first three hours of the day (9:00am to passed 12:00pm)
    t <- t + rexp(1, lambda_morn) # time lapsed while waiting for customer
    c <- c+1
  }
  while(t < 480){ # total number of minutes in day
    t <- t + rexp(1, lambda_aft) # time lapsed while waiting for customer
    c <- c+1
  }
  results[i] <- c
}
```

```{r q2_plot_results}
hist(results, 
     main="Customers each day",
     xlab="Customers",
     ylab="Days with that many customers"
)
```
```{r q2_mean_range}
print(paste0("Mean: ", round(mean(results), 2)))
print(paste0("Range: (", paste(range(results), collapse=", "), ")"))
```

## Observations

While the distribution still somewhat resembles an exponential distribution (notice the longer tail at the upper end), it also is not far from a Normal distribution. Many days had between 90 and 100 customers, and the mean is 97.57. Variation at the lower end is slightly smaller than variation at the upper end, with the day wit the least customers being ~30 customers less than the mean, and the day with the most customers being >40 more than the mean.

# Question 3