---
title: "DATA472 Assignment 3"
author: 'Loren Whithair (Student ID: 300 434 521)'
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  attr.source = ".numberLines")
```

# Question 1

## Code, with comments to identify randomness

Any line of code that adds randomness has been tagged as `#RANDOM`, with an additional tag of `INDEPENDENT` if the simulated values are independent of each other, and `DEPENDENT` otherwise.

First, initialise the variables for simulation and the population:
```{r q1_init_vars}
set.seed(42)

N <- 50 # population
L <- 100 # genetic markers
num_gen <- 20 # number of generations we will breed for
p <- 0.6 # initial probability of a "0" variant
```

Then, initialise a population as a matrix:
```{r q1_init_pop}
# get enough markers for the whole population
data <- sample(0:1, N*L*2, replace=TRUE, prob=c(p,1-p)) #RANDOM INDEPENDENT

# population matrix (one row per individual, one column per copy of a marker)
df <- matrix(data, nrow=N)
```

Create a function that breeds a new generation from the existing population:
```{r q1_breed_fn}
breed_new_gen <- function(df) {
  df_new <- df # work on a copy, so we always have initial population
  
  for (i in 1:nrow(df)){ # for every individual in the population...
    parents <- sample(1:nrow(df), 2, replace=FALSE) #RANDOM DEPENDENT
    
    for (j in 1:L) {# for every marker copy...
      # pick an index of which copy we will take from parent 1 and parent 2
      allele1 <- sample(1:2, 1, replace=TRUE) #RANDOM INDEPENDENT
      allele2 <- sample(1:2, 1, replace=TRUE) #RANDOM INDEPENDENT
      
      # get the variants from the parents
      copy1 <- df[parents[1], 2*(j-1)+ allele1]
      copy2 <- df[parents[2], 2*(j-1)+ allele2]
      
      # insert the parent's marker copies into the child's
      df_new[i,2*(j-1)+1] <- copy1
      df_new[i,2*(j-1)+2] <- copy2
    }
  }
  df_new
}
```

Using these components to breed for the specified number of generations:
```{r q1_breed}
df_new <- df
all_freqs <- matrix(NA, nrow=num_gen, ncol=L) #store the frequencies

for (g in 1:num_gen) {
  df_new <- breed_new_gen(df_new) # one round of breeding
  
  # save frequencies of this generation in row g. Each column is a 
  all_freqs[g,] <- sapply(1:L, function(j) {
    cols <- 2*(j-1) + 1:2
    sum(df_new[,cols] == 0)/N/2
  })
}
```

We now have `df` as the original generation and `df_new` as the final generation (50th).

```{r q1_sample_freq}

# get indices of individuals to survey from  1st and final population
s <- 20 # sample size
p1_idxs <- sample(1:N, s, replace=FALSE) #RANDOM DEPENDENT
pf_idxs <- sample(1:N, s, replace=FALSE) #RANDOM DEPENDENT

# get the individuals
df1 <- df[p1_idxs,]
dff <- df_new[pf_idxs,]

# calculate frequencies for histogram
freqs1 <- sapply(1:L, function(j) {
  cols <- 2*(j-1) + 1:2
  sum(df1[,cols] == 0)/s/2
})

freqsf <- sapply(1:L, function(j) {
  cols <- 2*(j-1) + 1:2
  sum(dff[,cols] == 0)/s/2
})

# generate histograms
xl <- c(0, 1)
yl <- c(0, 30)
par(mfrow=c(2,1))
hist(freqs1, 
     main="Frequencies of allele 0 in sample of first generation, across markers",
     xlab="Frequency of allele 0",
     ylab="Markers with this frequency",
     xlim=xl, ylim=yl
)
hist(freqsf, 
     main="Frequencies of allele 0 in sample of final generation, across markers",
     xlab="Frequency of allele 0",
     ylab="Markers with this frequency",
     xlim=xl, ylim=yl
)
```

## Observations of frequency

Both histograms are centered around approximately  `0.6`, the initial probability of a "0" variant.

The difference in variance demonstrates **genetic drift** - we can see that for the first generation, the frequency of a "0" for a given marker has 40% minimum, 80% maximum. For the final generation, there are at least some markers that consist entirely of the "0" variant.


